Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance

### 2019-05-05 Manuscript/manuscript.Rmd
manuscript edits

**Background.** Chromosome conformation capture sequencing technologies have shown that the three-dimensional (3D) structure of the genome is folded into distinct compartments, known as topologically associated domains (TADs) - units of coordinated gene expression. The location of TAD boundaries is highly conserved, suggesting the presence of epigenomic marks aiding in TAD formation. [The ability to predict which epigenomic features are most associated with the TAD boundaries will allow for better understanding the regulatory role of the 3D structure of the genome. - ??? Weak]

Existing methods for predicting associations between genomic annotations tend to ignore key characteristics of genomic data. Specifically, when constructing the model for classification, the number of genomic regions surrounding TAD boundaries is much less than the number of other regions. This leads to heavily imbalanced classes and can result in poor performances as a model's ability to assign a new sample to a given class is influenced by the prevalence of that class in the training set. Furthermore, most methods utilize direct overlap as a means to quantify this association, while distance, the measure of spatial relationships, remains unaccounted for.

**Methods.** We proposed a novel ensemble framework that systematically addresses those data irregularities and characteristics. Multiple random forest classification models, in combination with variations in cell line, data resolution, feature engineering, and re-sampling techniques were performed and compared. A number of performance metrics were assessed, including [??? will update when ready].

**Results.**

**Conclusions.**

@@

In this study, we have developed an ensemble framework for predicting TAD boundaries in order to systematically compare many data driven solutions used to address the issues presented above. We evaluated Hi-C data for both the GM12878 and K562 cell line obtained at 10 kb, 25 kb, 50 kb, and 100 kb resolution. For each resolution, we examined the inclusion of three different predictor types, including distance as the measure of association between genomic elements. We then tested four different re-sampling techniques aimed at creating balanced classes. Noninformative features were removed using a feature reduction algorithm and random forest classifications were performed to assess model performance, and investigate which genomic annotations were most associated with the TAD boundaries for a given system from the complete ensemble.  Model performance was evaluated using AUROC, AUPRC, F1-Score, and MCC.

>>>>

**Background.** Chromosome conformation capture sequencing technologies have shown that the three-dimensional (3D) structure of the genome is folded into distinct compartments, known as topologically associated domains (TADs) -- units of coordinated gene expression. The location of TAD boundaries has been shown to be highly conserved. Additionally, the distinct patterns of epigenomic marks, binding peaks of different factors, and actively transcribed genes prompt the use of computational approaches to identify the epigenomic features that are most predictive of TAD boundaries.

Current methods rely on the use of classification algorithms to draw associations between the spatial relationship of functional genomic elements and regions that include TAD boundaries. However, many methods ignore key characteristics of domain data, such as the impact of heavily imbalanced classes of boundary vs. non-boundary regions and the use of distances ($log_{2}$ transformed) as the best representation of the association between TAD boundaries and genomic annotations.

**Methods.** Using replicate Hi-C data for the GM12878 and K562 cell lines, we proposed a novel ensemble framework that addresses these data irregularities and characteristics. Multiple random forest classification models built on regularized sets of genomic annotations, in combination with variations in data resolution, feature engineering, and re-sampling techniques. The composite metric Matthew's Correlation Coefficient (MCC) was used to compare models derived from the framework to determine a consensus approach to predicting TAD boundary formation.

**Results.** We discovered that models built on SMOTE re-sampled data out performed models built simply on  random under- and over-sampled data. Additionally, our results indicated that $log_{2}$ distances between annotations and TAD boundaries were more informative than other predictor-types used to build models. Furthermore, we found that performances tended to degrade as data resolution decreased. Lastly, the most predictive annotations in the best performing models included known architectural proteins such as CTCF, SMC3, RAD21, and ZNF143, as well as lesser known transcription factors including P300 and NFKB.

**Conclusions.** The experiments derived from our ensemble framework provide a useful guide toward the exploration of chromatin organization, while highlighting the impact of both class imbalance and feature engineering in TAD boundary prediction.

@@

Therefore, in this study, we aimed to developed an ensemble framework used to build random forest classifiers composed of different combinations of data driven parameters used to address the issues presented above. We used our framework to systematically compare model performances using a composite performance metric, MCC. Our goal was to determine a suitable combination of data re-sampling and feature engineering techniques that both corrected the inherent imbalanced classes and provided the best spatial representation linking genomic annotations to TAD boundaries. Additionally, we used our framework to directly compare the influence of Hi-C contact matrix resolution on model performances. We evaluated Hi-C data for both the GM12878 and K562 cell line obtained at 10 kb, 25 kb, 50 kb, and 100 kb resolution. For each resolution, we examined the inclusion of three different predictor types, including distance as the measure of association between genomic elements. We then tested four different re-sampling techniques aimed at creating balanced classes. Noninformative features were removed using a feature reduction algorithm and random forest classifications were performed to assess model performance, and investigate which genomic annotations were most associated with the TAD boundaries for a given system from the complete ensemble.
